Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-07-01 content.tex
Fix URL
Spellcheck (there were no spelling errors found)
Updates from Overleaf

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
Bayesian phylogenetic inference often requires that the prior probability of the phylogeny according to the diversification model has to be computed millions of times. Therefore, biologically interesting but computationally expensive tree priors are non-viable. This is not necessarily problematic, when the data are very informative, as this will reduce the influence of the tree prior.
However, the assumption that tree prior choice is of low importance must first be verified.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior' $\mathit{p_{G}}$.
Many tree priors have a parameter setting for which it reduces to a standard tree prior. For example, a protracted birth-death model reduces to a standard birth-death model when speciation-completion rate approaches infinity [i.e. rate $\lambda$ in \cite{etienne2014estimating}, eqs.~(2b) and (2c)].
When benchmarking a novel tree prior, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations when assessing a novel tree prior, our examples contain only one replicate as they are for illustrative purposes only.

@@

Although many possible tests can be performed, 
we show the usage of \verb;pirouette; by a story that introduces
its features gradually, yet ending in quantifying the impact a tree prior has
in Bayesian inference.

@@

We assume the user has a phylogeny simulated with the new diversification model. The pipeline to assess the error BEAST2 makes in inferring this phylogeny then contains the following steps:
\begin{enumerate}
  \item from the given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is picked (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

@@

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree prior, clock model and site model) 
should include the generative/true tree prior. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when being the best.
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 4a),
and the inference models that were selected in the previous step (3). 
For each selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package which makes use of BEAST2. 
This step usually takes up most of the full pipeline's computation time.

The fourth step quantifies the inference error made. First the burn-in fraction is removed, i.e. the first phase of the Markov chain Monte Carlo (MCMC) run,
in which it samples an unrepresentative part of state space. By default, \verb;pirouette; 
removes the first 10\% of the posterior.
From the remaining posterior \verb;pirouette; 
creates an error distribution, by measuring the difference
between the true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The default way to quantify the difference between two phylogenies
is the nLTT statistic (\cite{janzen2015approximate}), but any 
user-defined error statistic can be used.

@@

From the twin tree, we simulate an alignment following the same
history as the twin tree, with the addition that 
the same total number of mutations have
taken place from root sequence to the sequences in the present.
This addition assures there is the same amount of historical information
stored in both the true and twin alignments.
We achieve this by simply simulating twin alignments until we
obtain one that has the desired number of mutations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative model only}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the second step we state our experiment.
We define an experiment $\mathit{X}$ as a combination of an inference model 
$\mathit{I}$ and conditions $\mathit{C}$.
In this example we pick $\mathit{I}$ to be the same inference model as 
the generative one,
which is the chosen Yule tree prior $\mathit{p_{G}^0}$ as well as site 
and clock models defined in $\mathit{A}$, respectively JC and strict.
We specify in $\mathit{C}$ that the experiment will always be run.

@@

We also need to specify the error measurement parameters $\mathit{E}$.
Here we use the default $\mathit{E}$, which uses a burn-in fraction 
of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

We now have all the needed \verb;pirouette; parameters: the alignment 
parameters, the experiments and the error measure parameters.
These objects need to be bundled in a bigger parameter structure, 
using \verb;create_pir_params;:

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it at section
 \ref{Comparing to background noise}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to other candidate models}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is 
non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be used in 
the inference. In this example, not only we re-use our hand-picked 
inference model, but we also pick the best inference model from a set of 
inference models. With the 'best' inference model, we mean the model with 
the highest evidence (also known as marginal likelihood), based on the 
alignment. We show the procedure using \verb;pirouette; to answer a second 
research question: "What is the inference error made on a novel phylogeny when using the best inference model, in comparison to a hand-picked model?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to background noise}\label{Comparing to background noise}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

So far we have measured the inference error on a tree
generated according to a known (and standard) tree prior. 
The goal of \verb;pirouette; is to measure the impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree prior, assumed to be closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the baseline 
error (that serves as a control) and the full inference error: it is their 
difference that shows the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for 
each of the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and a best candidate model, compared to the 
background noise?"

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_3}.

@@

We showed how to use \verb;pirouette; to quantify the importance of a 
tree prior in Bayesian phylogenetics, using the simplest generative tree 
prior possible.
In principle any other (more complex) generative tree prior can be tested, 
but we chose to provide the simplest (and fastest to run) examples.
All examples used only one original tree,
where any speciation process produces a whole range of trees.
One tree is not enough to determine the impact of a tree prior 
on Bayesian inference.
However, if the same procedure were repeated and performed on a 
distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the 
quality of the inference.
Also a twin tree does not always result in a lower error distribution,
as the stochasticity in generating a twin tree will occasionally
yield a tree of very low likelihood.

In conclusion, \verb;pirouette; can show the errors
when adopting a new generative prior
in comparison to the errors made by a standard prior.
From such results an user can estimate whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this were indeed the case, one could rightly implement 
their novel tree prior as an addition to 
their favorite Bayesian inference tool.

@@

The code used in this article and more examples that are periodically 
tested to work, can be found at 
\url{https://github.com/richelbilderbeek/pirouette}.

>>>>

Even when a diversification model passes the procedure described above,
it is not necessarily used in Bayesian inference.
Bayesian phylogenetic inference often requires that the prior probability of the phylogeny according to the diversification model has to be computed millions of times. Therefore, biologically interesting but computationally expensive tree priors are often not implemented, and simpler priors are used instead. This is not necessarily problematic, when the data are very informative, as this will reduce the influence of the tree prior.
However, the assumption that tree prior choice is of low importance must first be verified.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior' $\mathit{p_{G}}$.
Many tree priors have a parameter setting for which it reduces to a standard tree prior. For example, a protracted birth-death model \citep{etienne2012prolonging} reduces to a standard birth-death model when the speciation-completion rate is infinite [i.e. rate $\lambda$ in \cite{etienne2014estimating}, eqs.~(2b) and (2c)].
When benchmarking a novel tree prior, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations when assessing a novel tree prior, our examples contain only one replicate as they are for illustrative purposes only.

@@

Although many possible tests can be performed, 
we show the usage of \verb;pirouette; by introducing 
its features gradually, yet ending in quantifying the impact a tree prior has
in Bayesian inference.

@@

We assume the user has a phylogeny simulated with the new diversification model. The pipeline to assess the error BEAST2 makes in inferring this phylogeny then contains the following steps:
\begin{enumerate}
  \item from the given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is chosen (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

@@

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree prior, clock model and site model) 
should include the generative/true tree prior. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when it is the best.
The 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence for an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posterior distributions,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 4a),
and the inference models that were selected in the previous step (3). 
For each selected experiment a posterior distribution is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package which makes use of BEAST2. 
This step usually takes up most of the pipeline's computation time.

The fourth step quantifies the inference error made. First the burn-in fraction is removed, i.e. the first phase of the Markov chain Monte Carlo (MCMC) run,
which samples an unrepresentative part of state space. By default, \verb;pirouette; 
removes the first 10\% of the posterior.
From the remaining posterior \verb;pirouette; 
creates an error distribution, by measuring the difference
between the true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The default way to quantify the difference between two phylogenies
is the nLTT statistic (\cite{janzen2015approximate}), but any 
user-defined error statistic can be used.

@@

From the twin tree, we simulate an alignment following the same
history as the twin tree, with the constraint that 
the same total number of mutations have
taken place.
This constraint ensures there is the same amount of historical information
stored in both the true and twin alignments.
We achieve this by simply simulating twin alignments until we
obtain one that has the desired number of mutations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The generative and inference models are equal and standard.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the second step we state our experiment.
We define an experiment $\mathit{X}$ as a combination of an inference model 
$\mathit{I}$ and conditions $\mathit{C}$.
In this example we choose $\mathit{I}$ to be the same inference model as 
the generative one,
i.e. the Yule tree prior $\mathit{p_{G}^0}$ and site 
and clock models defined in $\mathit{A}$, respectively Jukes-Cantor and strict clock.
We specify in $\mathit{C}$ that the experiment will always be run.

@@

We also need to specify the error measurement parameters $\mathit{E}$.
Here we choose the default $\mathit{E}$, which has a burn-in fraction 
of 10\% and uses the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

We now have all the needed \verb;pirouette; parameters: the alignment 
parameters, the experiments and the error measure parameters.
These objects need to be bundled in a larger parameter structure, 
using \verb;create_pir_params;:

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it in section
 \ref{Comparing to background noise}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The inference model may differ from the generative model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$ is by (our) definition
non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be used in 
the inference. In this example, we do not only re-use our hand-picked 
inference model, but we also pick the best inference model from a set of 
inference models, i.e.  the model with 
the highest evidence measured by its marginal likelihood. We show the procedure using \verb;pirouette; to answer a second 
research question: "What is the inference error made on a novel phylogeny when using the best inference model, in comparison to a hand-picked model?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The generative model is non-standard}\label{Comparing to background noise}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

So far we have measured the inference error on a tree
generated according to a known (and standard) tree prior. 
The goal of \verb;pirouette; is to measure the impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree prior, assumed to be closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the baseline 
error (that serves as a control) and the full inference error: it is their 
difference that shows the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for 
both the generative and best candidate models.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and the best candidate model?"

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_3}. While the error distributions using the best or generative model as inference model are very similar, the error distributions of the true tree are substantially larger than those of the twin tree. This is the error made in inference by not using the generating species tree prior.

@@

We showed how to use \verb;pirouette; to quantify the importance of a 
tree prior in Bayesian phylogenetics, using the simplest generative tree 
prior possible.
In principle any other (more complex) generative tree prior can be tested, 
but we chose to provide the simplest (and fastest to run) examples.

Figure~\ref{fig:example_3} illustrates the primary result of our pipeline: it shows the error distributions for the true tree and the twin tree when either the generating model or the best candidate model is used. For this specific case it does not seem to matter that the site and clock models used in inference are different from those used in generating the data, but the difference between the error distributions for the true tree and the twin tree suggest that the choice of tree prior does matter.

We note, however, that all examples used only one original tree,
where any speciation process produces a whole range of trees.
One tree is not enough to determine the impact of a tree prior on Bayesian inference.
However, if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the 
quality of the inference.
Also a twin tree does not always result in a lower error distribution,
as the stochasticity in generating a twin tree will occasionally
yield a tree of very low probability.

In conclusion, \verb;pirouette; can show the errors
when adopting a new generative prior
in comparison to the errors made by a standard prior.
The user can then judge whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this is indeed the case, one should implement 
the novel tree prior as an addition to 
the favorite Bayesian inference tool.

@@

The code used in this article and more examples that are periodically 
tested, can be found at 
\url{https://github.com/richelbilderbeek/pirouette_examples}.
